---
title: "Keras/TensorFlow installation and configuration"
author: "Diego Mañanes"
date: "`r paste0(Sys.Date())`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
geometry: margin=3cm
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Installation

_digitalDLSorteR_ is based on Deep Neural Networks (DNN) models. To use them, it uses `r CRANpkg("keras")` package (available from CRAN), a high-level neural networks API implemented in Python. `r CRANpkg("keras")` R version works as an interface between these two languages by using `r CRANpkg("reticulate")` package. Therefore, `r CRANpkg("keras")` requires a Python interpreter to work. We recommend using conda environments to provide a Python interpreter with all its dependencies covered. If you have a conda environment compatible with `r CRANpkg("keras")` requirements, the package will find it automatically during the installation. Otherwise, `r CRANpkg("keras")` will create a new environment named `r-reticulate` with all these dependencies covered. There are other methods to install a functionally back-end, see <https://keras.rstudio.com/reference/install_keras.html> for more details.

```{r, eval = FALSE}
install.packages("keras")
library(keras)
install_keras(method = "conda") # other options are available
```

This is the default CPU-based installation, but it is possible a more customized installation. See <https://tensorflow.rstudio.com/installation/> and <https://keras.rstudio.com/reference/install_keras.html> for more details.

## Common errors in Keras

### Keras package does not find any functional Python environment

If you experiment errors related to its installation and/or `r CRANpkg("keras")` is not able to find a functional Python environment, follow these steps:

Instead of using `keras::install_keras()` function, run the following code chunk installing `r CRANpkg("tensorflow")` R package. This code will create a Python interpreter with `r CRANpkg("tensorflow")` with all its dependencies covered.

```{r, eval = FALSE}
tensorflow::install_tensorflow(
  method = 'conda', envname = 'r-reticulate-tensorflow', version = '2.1.0-cpu'
)
```

If this code chunk raises an error, create manually a conda environment from any terminal:

```{bash}
conda create --name r-tensorflow-21 python tensorflow=2.1
```

Then, run the above chunk code setting the `envname` parameter equal to the name of the newly created conda environment name. 

Finally, problems still happen, set this new environment as the selected Python environment:

```{r}
tensorflow::use_condaenv("r-reticulate-tensorflow")
```

### Eager execution

Depending on `r CRANpkg("tensorflow")` version used as back-end, you are likely to experiment an error related to eager execution. Eager execution is an imperative, define-by-run interface where operations are executed immediately as they are called. However, it is only available on `r CRANpkg("tensorflow")` versions later than 1.9. You can disable this option by the following line of code or change your `r CRANpkg("tensorflow")` version used as back-end.

```{r}
tensorflow::tf$compat$v1$disable_eager_execution()
```
