---
title: "Keras/TensorFlow installation and configuration"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    vignette: >
      %\VignetteIndexEntry{Keras/TensorFlow installation and configuration}
      %\VignetteEngine{knitr::rmarkdown}
      %\VignetteEncoding{UTF-8}
bibliography: references.bib
geometry: margin=3cm
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Installation

**digitalDLSorteR** is based on Deep Neural Networks (DNN) models. To use them, [keras](https://cran.r-project.org/package=keras) package (available from CRAN), a high-level neural networks API implemented in Python, is used. [keras](https://cran.r-project.org/package=keras) R version works as an interface between these two languages by using [reticulate](https://cran.r-project.org/package=reticulate) package. Therefore, [keras](https://cran.r-project.org/package=keras) requires a Python interpreter to work. We recommend using conda environments to provide a Python interpreter with all its dependencies covered. If you already have a conda environment compatible with [keras](https://cran.r-project.org/package=keras) requirements, the package will find it automatically during the installation. Otherwise, [keras](https://cran.r-project.org/package=keras) will create a new environment named `r-reticulate` with all these dependencies covered. There are other methods to install a functionally back-end and it is possible more customized installations, see <https://keras.rstudio.com/reference/install_keras.html> and <https://tensorflow.rstudio.com/installation/> for more details.

```{r, eval = FALSE}
install.packages("keras")
library(keras)
install_keras(method = "conda") # other options are available
```


## Common errors with [keras](https://cran.r-project.org/package=keras)

### [keras](https://cran.r-project.org/package=keras) package does not find any functional Python environment

If you experiment errors related to its installation and/or [keras](https://cran.r-project.org/package=keras) is not able to find a functional Python environment, follow these steps:

Instead of using `keras::install_keras()` function, run the following code chunk installing [tensorflow](https://cran.r-project.org/package=tensorflow) R package. This code will create a Python interpreter with [tensorflow](https://cran.r-project.org/package=tensorflow) with all its dependencies covered.

```{r, eval = FALSE}
tensorflow::install_tensorflow(
  method = 'conda', envname = 'r-reticulate-tensorflow', version = '2.1.0-cpu'
)
```

If this code chunk raises an error, create manually a conda environment from any terminal:

```{bash}
conda create --name r-tensorflow-21 python tensorflow=2.1
```

Then, run the above chunk code setting the `envname` parameter equal to the name of the newly created conda environment name. If problems still happen, set this new environment as the selected Python environment:

```{r}
tensorflow::use_condaenv("r-reticulate-tensorflow")
```

Finally, if Python-interpreter problems don't dissapear, it is possible to set manually the interpreter that will be used using [reticulate](https://cran.r-project.org/package=reticulate). See the following guide: <https://rstudio.github.io/reticulate/articles/versions.html>. And of course, don't hesitate to tell us your problem and we'll try to help you!

### Eager execution

Depending on [tensorflow](https://cran.r-project.org/package=tensorflow) version used as back-end, you are likely to experiment an error related to eager execution. Eager execution is an imperative, define-by-run interface where operations are executed immediately as they are called. However, it is only available in [tensorflow](https://cran.r-project.org/package=tensorflow) versions later than 1.9. You can disable this option by the following line of code or change your [tensorflow](https://cran.r-project.org/package=tensorflow) version used as back-end.

```{r}
tensorflow::tf$compat$v1$disable_eager_execution()
```
