---
title: "Using pre-trained context-specific deconvolution models"
author: "Diego Mañanes"
date: "`r paste0(Sys.Date())`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
geometry: margin=3cm
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using pre-trained models

_digitalDLSorteR_ offers the possibility of using pre-trained context-specific deconvolution models included in _digitalDLSorteRdata_ (<https://github.com/diegommcc/digitalDLSorteRdata>) to deconvolve new bulk RNA-Seq samples from the same biological environment. This is the simplest way to use _digitalDLSorteR_ and only requires to load a raw bulk RNA-Seq matrix with genes as rows (annotated as _SYMBOL_) and samples as columns, and select the wanted model. It is done using `deconvDigitalDLSorter` function that by default normalize the samples to counts per million (CPMs), so this matrix should be raw counts.

**Figure 1**

## Available models

So far, the available models just cover two possible biological environments: breast cancer and colorectal cancer. These models are able to accurately deconvolve new samples from their same environment as they have been trained with transcriptional profiles from these specific contexts. 

### Breast cancer models

There are two different deconvolution models for breast cancer samples differing in the level of specificity. Both of them have been trained using data from [Chung et al., 2017](https://www.nature.com/articles/ncomms15081) ([GSE75688](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE75688)).

* `breast.chung.generic`: this model considers 13 cell types, four of them being the intrinsic molecular subtypes of breast cancer (`ER+`, `HER2+`, `ER+/HER2+` and `TNBC`) and the rest immune and stromal cells (`Stromal`, `Monocyte`, `TCD4mem` (memory CD4+ T cells), `BGC` (germinal center B cells), `Bmem` (memory B cells), `DC` (dendritic cells), `Macrophage`, `TCD8` (CD8+ T cells) and `TCD4reg` (regulatory CD4+ T cells)).
* `breast.chung.generic`: this model considers 7 cell types. They are generic groups of the cell types considered by the specific version: B cells (`Bcell`), T CD4+ cells (`TcellCD4`), T CD8+ cells (`TcellCD8`), monocytes (`Monocyte`), dendritic cells (`DCs`), stromal cells (`Stromal`) and tumor cells (`Tumor`).

### Colorectal cancer model

This model was created with data from [Li et al., 2017](https://www.nature.com/articles/ng.3818) ([GSE81861](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE81861)). It is called `colorectal.li'  and considers 10 cell types: cancer cells (`CRC`), epithelial (`Ep`), monocytes (`M`), macrophages (`Mc`), fibroblasts (`Fb`), CD4 T cells (`CD4`), CD8 T cells (`CD8Gp`), CD8 T cells (`CD8Gn`), germinal center B cells (`gB`) and mature B cells (`pB`).

## Example using TCGA breast cancer samples

The following chunk shows an example using `breast.chung.generic` model and a small portion of TCGA data loaded from `digitalDLSorteRdata`:

```{r deconvolvingTCGA, message=FALSE}
library(digitalDLSorteR)
library(digitalDLSorteRdata)
# disable eager execution, compatibility issues
tensorflow::tf$compat$v1$disable_eager_execution()
# loading model and example data from digitalDLSorteRdata
data(breast.chung.generic)
data(TCGA.breast.small)
```

`breast.chung.generic` is a `DigitalDLSorterDNN` object which can be inspected in order to check which cell types are considered by the model, the performance that it had on test data, etc.

```{r}
breast.chung.generic
```

```{r}
cell.types(breast.chung.generic)
```

Now, we can use it to deconvolve `TCGA.breast.small` samples as follows:

```{r}
# deconvolution
deconvResults <- deconvDigitalDLSorter(
  data = TCGA.breast.small,
  model = breast.chung.generic,
  normalize = TRUE
)
rownames(deconvResults) <- paste("Sample", seq(nrow(deconvResults)), sep = "_")
head(deconvResults)
```


`deconvDigitalDLSorter` returns a data frame with samples as rows ($k$) and cell types considered by the model as columns ($j$). Each entry corresponds with the proportion of $k$ cell type in $i$ sample. In order to evaluate visually these results by a bar plot, you can use `barplotCellTypes` function as follows:

```{r resultsDeconvTCGA}
barPlotCellTypes(
  deconvResults, 
  color.line = "black",
  title = "Results of deconvolution of TCGA breast samples"
)
```

`deconvDigitalDLSorterObj` also offers some parameters that can be used in the case of simplify your results by adding up cell proportions from similar cell types: `simplify.set` and simplify.majority`. For instance, we can summarize dendritic cells (DCs) and monocytes into mononuclear phagocyte system (MPS) using `simplify.set` as follows:

```{r}
# deconvolution
deconvResultsSum <- deconvDigitalDLSorter(
  data = TCGA.breast.small,
  model = breast.chung.generic,
  normalize = TRUE,
  simplify.set = list(MPS = c("DCs", "Monocyte"))
)
rownames(deconvResultsSum) <- paste("Sample", seq(nrow(deconvResults)), sep = "_")
barPlotCellTypes(
  deconvResultsSum, 
  rm.x.text = FALSE,
  color.line = "black",
  title = "Results of deconvolution of TCGA breast samples (simplified)"
)
```

On the other hand, `simplify.majority` does not create new classes but adds up the proportions of the most abundant cell type in each sample from those provided. Look at the documentation and examples for more details.


## Contribute with your own models

New models are thought to be published coming soon, but you can contribute with your own models making them available by contacting us.

## References



## Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

