---
title: "Performance of a real model: deconvolution of colorectal cancer samples"
author: "Diego Mañanes"
date: "`r paste0(Sys.Date())`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
geometry: margin=3cm
fontsize: 12pt
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Performance of a real model: deconvolution of colorectal cancer samples

In "Building new deconvolution models" was shown the workflow needed to build new deconvolution models. However, it weas done using a 'toy' example in order to avoid long execution times. In this vignette, the performance of the real model trained with the full dataset is shown to provide a guide for new users that want to build new models. 

## Loading and inspection of the model

Model is going to be loaded from _digitalDLSorteRdata_ package. It is a `DigitalDLSorter` object with `prob.cell.matrix` and `trained.model` slots:

```{r}
library(digitalDLSorteRdata)
library(digitalDLSorteR)
data(DDLSLiComp)

DDLSLiComp
```

As it is shown above, the model was trained 30 epochs with a total of 20,000 pseudo-bulk samples (13,334 for training and 6,666 for test). Now, evaluation metrics are going to be calculated by `calculateEvalMetrics` function.

```{r}
DDLSLiComp <- calculateEvalMetrics(DDLSLiComp)
```

And then, evaluation plots can be displayed:

```{r distErr1}
distErrorPlot(
  DDLSLiComp,
  error = "AbsErr",
  x.by = "CellType",
  color.by = "CellType", 
  error.labels = FALSE, 
  type = "boxplot",
  size.point = 0.5
)
```


```{r distErr2}
distErrorPlot(
  DDLSLiComp,
  x.by = "pBin",
  error = "AbsErr",
  facet.by = "CellType",
  color.by = "nCellTypes", 
  type = "violinplot",
  size.point = 0.5
)
```


```{r distErr3}
distErrorPlot(
  DDLSLiComp,
  error = "AbsErr",
  color.by = "CellType", 
  facet.by = "nCellTypes",
  type = "boxplot",
  size.point = 1
)s
```


```{r barError}
barErrorPlot(DDLSLiComp, error = "MAE", by = "CellType")
```





```{r corr1}
corrExpPredPlot(
  DDLSLiComp,
  color.by = "CellType",
  size.point = 1,
  corr = "both"
)
```


```{r corr2}
corrExpPredPlot(
  DDLSLiComp,
  color.by = "CellType",
  facet.by = "CellType",
  size.point = 0.5, 
  filter.sc = F,
  corr = "both"
)
```

```{r corr3}
corrExpPredPlot(
  DDLSLiComp,
  color.by = "CellType",
  facet.by = "nCellTypes",
  size.point = 1,
  corr = "both"
)
```




```{r bland1}
blandAltmanLehPlot(
  DDLSLiComp, 
  color.by = "CellType",
  log.2 = FALSE,
  size.point = 1,
  filter.sc = TRUE,
  density = TRUE,
)
```


```{r bland2}
blandAltmanLehPlot(
  DDLSLiComp, 
  color.by = "nCellTypes",
  facet.by = "nCellTypes",
  log.2 = FALSE,
  size.point = 1,
  filter.sc = TRUE,
  density = TRUE
)
```
